Synthesizing Quotient Lenses

### 2017-10-05 papers/QuotientLensSynth/main.tex
Bunch of comments here and there

\saz{It might be a good idea to show what this would look like in all its gory
detail, just to make the problem very concrete. Perhaps in a figure?}

@@

Our plan for this paper is as follows:
\begin{enumerate}
  \item
  We introduce the language of Quotient Regular Expressions (QREs) which enables
  us to express a set of equivalence relations on regular languages using a
  compact, easy-to-use syntax.
  \item
  We describe a class $\mathcal{Q}$ of quotient lenses whose types are given by
  QREs. We derive a normal form for quotient lenses in $\mathcal{Q}$, and
  demonstrate that elements of $\mathcal{Q}$ that are in this normal form are
  closed under left quotienting, right quotienting, composition and regular
  operators.
  \item
  Using the normal form for quotient lenses in $\mathcal{Q}$, we reduce the
  synthesis of quotient lenses in $\mathcal{Q}$ to the synthesis of bijective
  lenses between regular languages. 
  \item
  We describe an implementation of QREs, the language $\mathcal{Q}$ of quotient
  lenses, and an algorithm for synthesizing quotient lenses in $\mathcal{Q}$ in
  the Boomerang language.
  \item
  We assess the usabibility of the QRE synthesis via a case study in which we
  selected several data formats used by data.gov and synthesized quotient lenses
  between them.
\end{enumerate}

@@

\sam{More explanation here?}

@@

Each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $EqRel(q)$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$, and
  \item a ``canonizing'' function $Canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $Canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $EqRel(q)$
  \end{enumerate}
  Observe that since $Canonizer(q)$ computes $Canonizer(q)(w)$ as the unique $k$
  in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $EqRel(q)$,
  then the equivalence classes of $EqRel(q)$ are the same as the fibres of
  $Canonizer(q)$.
  \subsection{Meaning and Example Usages for QREs}
  \begin{enumerate}
    \item 
    Every regular expression is a QRE. When a regular expression $R$ is used as
    a QRE, then $\mathcal{L}(R)$ is assumed to be quotiented by the equality
    equivalence relation on $\mathcal{L}(R)$.
    \item
    For the QRE $R \mapsto s$, the data is described by $\mathcal{L}(R)$, but
    all members of $R$ are considered equivalent to each other, with $s \in
    \mathcal{L}(R)$ as the choice of representative. This construct is useful
    when the programmer simply wishes to ignore unnecessary characters such as
    excess whitespace e.g. 
    $$FIRST\_NAME \cdot (WHITESPACE^* \mapsto `` \quad ") \cdot LAST\_NAME$$
    \item
    $\squash{R}{R'}{f}$ enables the programmer to use one of two different
    representations of the data if they can provide a way of mapping the first
    representation to the second. For instance, the first representation may be
    $$FIRST\_NAME \cdot `` \quad " \cdot LAST\_NAME$$
    while the second representation may be
    $$FIRST\_NAME \cdot ``," \cdot LAST\_NAME$$
    The mapping that maps the first representation to the second simply converts
    the space between the first and last name to a comma.
    \item
    The $\perm{q_1, \ldots, q_n}{q}$ construct enables the programmer to
    consider all permutations of some data that are interspersed with a
    separator the same. For instance, in the \textsc{Bib}\TeX{} to EndNote
    example, the programmer may use this construct as follows:
    $$\perm{\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},
    \re{YEAR}}{(\re{WSP}^* \mapsto \string \n)}$$
 where $\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},$ and $\re{YEAR}$ are
regular expressions describing the respective \textsc{Bib}\TeX\; fields of the
same name, and $WSP$ is a regular expression describing whitespace characters.
\item
The constructs $q \cdot q'$, $q \sep q'$ and $q^*$ are similar to the respective
regular expression constructs. However, the regular constructs for QREs are more
restrictive for QREs since the data should match the whole language $W(q)$ and
the kernel language $K(q)$ unambiguosly so as to make transformations
deterministic.
\item
The $\normalize{R}{R'}{f}$ construct allows the programmer to choose how to
canonize $\mathcal{L}(R)$ into $\mathcal{L}(R')$ be providing the canonizing
function $f$ themselves. This constuct is useful when the desired equivalence
relation cannot easily be expressed by any of the other constructs.
  \end{enumerate}

@@

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      R \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q^* & W(q)^* & K(q)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{(q_1, \ldots, q_n)}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot \cdot W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{(q_1, \ldots, q_n)}{q} ) = &
 K(q_1) \cdot K(q_1) \cdot \ldots \cdot \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions.}
  \label{fig:wk}
\end{figure}

@@

    $w = r_{\sigma(1)} \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}
    \text{ and } w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
    \cdot {r'}_{\theta(n)}$ for some $\sigma, \theta \in S_n$, with $r_i \;
    \equiv_{q_i} \; r'_i$ and $s_k \; \equiv_{q} \; s'_{k}$.
  \caption{QRE Equivalence Relations}
  \label{fig:relations}
\end{figure}
\begin{figure}[t]
  \begin{center}
\[
    \begin{array}{l@{\quad}l} 
      q & Canonizer(q)  \\ \hline
      R & Canonizer(R) = id_{\mathcal{L}(R)} \\
      R \mapsto s & Canonizer(R \mapsto s)(w) = s\\
      \squash{R}{R'}{f} & Canonizer(\squash{R}{R'}{f})(w) = 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}\\
      \normalize{R}{R'}{f} & Canonizer(\normalize{R}{R'}{f}) = f\\
      \perm{(q_1, \ldots, q_n)}{q} & \\
      q_2 \circ  q_1 & Canonizer(q' \circ q) = Canonizer(q') \circ Canonizer(q)\\
      q_1 \cdot q_2 & Canonizer(q' \cdot q) = Canonizer(q') \cdot Canonizer(q)\\
      q_1 \sep q_2 &  Canonizer(q_1 \sep q_2)(w) = 
\begin{cases}
Canonizer(q_1)(w) & \text{if } w \in \mathcal{L}(W(q_1))\\
Canonizer(q_2)(w) & \text{if } w \in \mathcal{L}(W(q_2))\\
\end{cases}\\
      q^* &  Canonizer(q^*) = Canonizer(q)^* \\
    \end{array}
    \]
    \end{center}
    $Canonizer(\perm{(q_1, \ldots, q_n)}{q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (Canonizer(q_1)(r_1)) \cdot (Canonizer(q)(s_1)) \cdot \ldots \cdot
(Canonizer(q_n)(r_n))$
  \caption{QRE Canonizers}
  \label{fig:canonizers}
\end{figure}
\begin{figure}[t]
\centering
\begin{prooftree}
\AxiomC{}
\UnaryInfC{$R$ is well formed}
\end{prooftree}

@@

  \caption{QRE Inference Rules}

@@

derive elements of $\mathcal{Q}$:

@@

\begin{claim}
If there exists a derivation $q:c \Leftrightarrow c'$ then
$\llbracket q \rrbracket$ is a quotient lens from $R$ to $S$.
\end{claim}

>>>>

\saz{It might be a good idea to show what this would look like in all its gory
detail, just to make the problem very concrete. Perhaps in a figure?}\bcp{+1}

@@

Our plan for this paper is as follows:\bcp{This should not just explain the
  plan, but rather state the contributions {\em as contributions},
  not only summarizing what they are but suggesting why they are novel,
  challenging, and important.}
\begin{enumerate}
  \item
  We introduce the language of Quotient Regular Expressions (QREs) which enables
  us to express a set of equivalence relations on regular languages using a
  compact, easy-to-use syntax.
  \item
  We describe a class $\mathcal{Q}$ of quotient lenses whose types are given by
  QREs. We derive a normal form for quotient lenses in $\mathcal{Q}$, and
  demonstrate that elements of $\mathcal{Q}$ that are in this normal form are
  closed under left quotienting, right quotienting, composition and regular
  operators.
  \item
  Using the normal form for quotient lenses in $\mathcal{Q}$, we reduce the
  synthesis of quotient lenses in $\mathcal{Q}$ to the synthesis of bijective
  lenses between regular languages. 
  \item
  We describe an implementation of QREs, the language $\mathcal{Q}$ of quotient
  lenses, and an algorithm for synthesizing quotient lenses in $\mathcal{Q}$ in
  the Boomerang language.
  \item
  We assess the usabibility of the QRE synthesis via a case study in which we
  selected several data formats used by data.gov and synthesized quotient lenses
  between them.
\end{enumerate}

@@

\sam{More explanation here?}\bcp{Probably not needed---we're going straight
  on to the detailed explanation anyway.}

@@

Each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $EqRel(q)$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$, and
  \item a ``canonizing'' function\bcp{Words in math mode need $\backslash${\tt
      mathit} around them, always.} $Canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $Canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $EqRel(q)$
  \end{enumerate}
  Observe that since $Canonizer(q)$ computes $Canonizer(q)(w)$ as the unique $k$
  in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $EqRel(q)$,
  then the equivalence classes of $EqRel(q)$ are the same as the fibres of
  $Canonizer(q)$.
  \subsection{Meaning and Example Usages for QREs}
  \begin{enumerate}
    \item 
    Every regular expression is a QRE. When a regular expression $R$ is used as
    a QRE, then $\mathcal{L}(R)$ is assumed to be quotiented by the equality
    equivalence relation on $\mathcal{L}(R)$.
    \item
    For the QRE $R \mapsto s$, the data is described by $\mathcal{L}(R)$, but
    all members of $R$ are considered equivalent to each other, with $s \in
    \mathcal{L}(R)$ as the choice of representative. This construct is useful
    when the programmer simply wishes to ignore unnecessary characters such as
    excess whitespace e.g. \bcp{Suggest using $\backlash${\tt mathtt} for 
      concrete program identifiers and ascii characters in examples.}
    $$FIRST\_NAME \cdot (WHITESPACE^* \mapsto `` \quad ") \cdot LAST\_NAME$$
    \item
    $\squash{R}{R'}{f}$ enables the programmer to use one of two different
    representations of the data if they can provide a way of mapping the first
    representation to the second. For instance, the first representation may be
    $$FIRST\_NAME \cdot `` \quad " \cdot LAST\_NAME$$
    while the second representation may be
    $$FIRST\_NAME \cdot ``," \cdot LAST\_NAME$$
    The mapping that maps the first representation to the second simply converts
    the space between the first and last name to a comma.
    \item
    The $\perm{q_1, \ldots, q_n}{q}$ construct enables the programmer to
    consider all permutations of some data that are interspersed with a
    separator the same. For instance, in the \textsc{Bib}\TeX{} to EndNote
    example, the programmer may use this construct as follows:
    $$\perm{\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},
    \re{YEAR}}{(\re{WSP}^* \mapsto \string \n)}$$
 where $\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},$ and $\re{YEAR}$ are
regular expressions describing the respective \textsc{Bib}\TeX\; fields of the
same name, and $WSP$ is a regular expression describing whitespace characters.
\item
The constructs $q \cdot q'$, $q \sep q'$ and $q^*$ are similar to the respective
regular expression constructs. However, the regular constructs for QREs are more
restrictive for QREs since the data should match the whole language $W(q)$ and
the kernel language $K(q)$ unambiguosly so as to make transformations
deterministic.
\item
The $\normalize{R}{R'}{f}$ construct allows the programmer to choose how to
canonize $\mathcal{L}(R)$ into $\mathcal{L}(R')$ be providing the canonizing
function $f$ themselves. This constuct is useful when the desired equivalence
relation cannot easily be expressed by any of the other constructs.
  \end{enumerate}

@@

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      R \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q^* & W(q)^* & K(q)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{(q_1, \ldots, q_n)}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot \cdot W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{(q_1, \ldots, q_n)}{q} ) = &
 K(q_1) \cdot K(q_1) \cdot \ldots \cdot \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions\bcp{left-justify columns}}
  \label{fig:wk}
\end{figure}

@@

    $w = r_{\sigma(1)} \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}
    \text{ and } w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
    \cdot {r'}_{\theta(n)}$ for some $\sigma, \theta \in S_n$, with $r_i \;
    \equiv_{q_i} \; r'_i$ and $s_k \; \equiv_{q} \; s'_{k}$.
  \caption{QRE Equivalence Relations\bcp{too wide}}
  \label{fig:relations}
\end{figure}
\begin{figure}[t]
  \begin{center}
\[
    \begin{array}{l@{\quad}l} 
      q & Canonizer(q)  \\ \hline
      R & Canonizer(R) = id_{\mathcal{L}(R)} \\
      R \mapsto s & Canonizer(R \mapsto s)(w) = s\\
      \squash{R}{R'}{f} & Canonizer(\squash{R}{R'}{f})(w) = 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}\\
      \normalize{R}{R'}{f} & Canonizer(\normalize{R}{R'}{f}) = f\\
      \perm{(q_1, \ldots, q_n)}{q} & \\
      q_2 \circ  q_1 & Canonizer(q' \circ q) = Canonizer(q') \circ Canonizer(q)\\
      q_1 \cdot q_2 & Canonizer(q' \cdot q) = Canonizer(q') \cdot Canonizer(q)\\
      q_1 \sep q_2 &  Canonizer(q_1 \sep q_2)(w) = 
\begin{cases}
Canonizer(q_1)(w) & \text{if } w \in \mathcal{L}(W(q_1))\\
Canonizer(q_2)(w) & \text{if } w \in \mathcal{L}(W(q_2))\\
\end{cases}\\
      q^* &  Canonizer(q^*) = Canonizer(q)^* \\
    \end{array}
    \]
    \end{center}
    $Canonizer(\perm{(q_1, \ldots, q_n)}{q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (Canonizer(q_1)(r_1)) \cdot (Canonizer(q)(s_1)) \cdot \ldots \cdot
(Canonizer(q_n)(r_n))$
  \caption{QRE Canonizers\bcp{First column not needed (and, once it's gone,
      maybe the = signs could be vertically aligned?)}}
  \label{fig:canonizers}
\end{figure}
\begin{figure}[t]
\centering
\begin{prooftree}
\AxiomC{}
\UnaryInfC{$R$ is well formed}
\end{prooftree}

@@

  \caption{QRE Inference Rules\bcp{Maybe some or all of these can be
      suppressed in the ``short version'' of the paper}}

@@

derive elements of $\mathcal{Q}$:\bcp{I think the numbering and the ``if $q
  = ..." stuff is not needed.  And maybe we can save space by typesetting the
  get/put parts on the same line as the typing rule, at least in some cases.}

@@

\begin{claim}
If there exists a derivation $q:c \Leftrightarrow c'$ then
$\llbracket q \rrbracket$ is a quotient lens from $R$ to $S$. \bcp{??}
\end{claim}
